---
title: "Using nls function for nonlinear regression"
subtitle: "*modeling cell survival curve after radiation*"
output: html_document
---
We can use the Poisson distribution to model radiation hit in target cells.
$$P(n)=\frac{\lambda^ne^{-\lambda}}{n!}$$
where $\lambda$ is the average number of events occurred in a unit time and n is the specific number of events occurred in a unit time. 

If each "hit" is assumed to result in cell inactivation, then the probability of survival is the probability of not being hit, P(0).
$$P(0)=\frac{\lambda^0e^{-\lambda}}{0!}=e^{-\lambda}$$
We define a dose $D_0$ that delivers, on average, one lethal event per target. This dose will result in $P(0)=e^{-1}=0.37$ survival. $D_0$ is often called the **mean lethal dose**.

#### Linear-Quadratic Model
The *linear-quadratic* (**L-Q**) equation is the most wildly accepted method of fitting the survival of cells following radiation. It is given by $$S=e^{-(\alpha D+\beta D^{2})}$$
Where S is the number of surviving cells following a dose of D, and $\alpha$ and $\beta$ describe the linear and quadratic parts of the survival curve. The $\alpha$ and $\beta$ constants vary between different tissues and tumors.

#### Multi-Target Model
In a *multi-target* (**M-T**) model, each cell contains n distinct and identical targets and all n targets must be inactivated to kill the cell. The probability that a target is not hit is $e^{-\frac{D}{D_0}}$. The probability that a target is hit is $1-e^{-\frac{D}{D_0}}$. The probability that all n targets are hit is $(1-e^{-\frac{D}{D_0}})^n$. Therefore the probability that all n targets will not be hit, i.e., the probability of survival, is $$S=1-(1-e^{-\frac{D}{D_0}})^n$$

Here this clonogenic assay shows the survival of FaDu cells following radiation and we are going to fit the dose response curve to the L-Q and M-T models.

Load and view data.

```{r load and view data, echo=FALSE, out.width="60%"}
data<-data.frame(radiation=c(0,2,4,6,8),surviving.fraction=c(0.992063492,0.523809524,0.170634921,0.037037037,0.005291005),sd=c(0.297253744,0.153566814,0.085736329,0.00819679,0.00819679))
plot(data$radiation,data$surviving.fraction,pch=19,xlab="Dose (Gy)",ylab="Surviving fraction")
```

Transforming the L-Q equation to $lnS=-\alpha D-\beta D^2$ reveals that we can use a polynomial regression to get starting estimates.

```{r poly model, results='hide'}
model_poly <- lm(-log(surviving.fraction)~-1+radiation+I(radiation^2), data=data) #-1 is to set the intercept to 0
summary(model_poly)
```

```{r nls lq model, results='hide'}
model_lq<- nls(surviving.fraction~exp(-a*radiation-b*radiation^2),data=data, start = list(a=coef(model_poly)[1],b=coef(model_poly)[2]))
summary(model_lq)
```

To fit the M-T model, I don't have a way of estimating the starting values. Try a few numbers to feed the model.

```{r nls mt model, results='hide'}
model_mt <- nls(surviving.fraction ~ 1-(1-exp(-radiation/D0))^n, data=data, start = list(D0=2,n=2))
summary(model_mt)
```

To create the fitted regression lines
```{r regression line}
# create more "radiation" points for the fitted line
newdata <- data.frame(radiation = rep(seq(0,8, length = 100),2), model=rep(c("lq","mt"), each=100))

# predict surviving fractions for each radiation dose based on each model
newdata$prediction <- c(predict(model_lq,newdata)[1:100], predict(model_mt,newdata)[1:100])
```

Final plot with ggplot2

```{r ggplot, echo=FALSE, warning=FALSE}
library(ggplot2)
ggplot(data=data, aes(x = radiation, y = surviving.fraction)) +
  geom_point(size = I(3)) +
  geom_errorbar(data=data, aes(ymin=surviving.fraction-sd, ymax=surviving.fraction+sd), width=.1) + # error bars
  geom_line(data=newdata,aes(x = radiation, y = prediction, color = model), size = I(1)) +
  xlab("Dose (Gy)") + ylab("Surviving fraction") +
  scale_color_discrete(labels=c("linear-quadratic","multi-target"))+ # change the labels in legend
  scale_y_log10() + # change y axis to log scale
  theme(legend.position = c(0.9,0.9),
        legend.justification = c(1,1),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.line = element_line(size = 1))
```

So the FaDu cell line has $\alpha=$ `r round(coef(model_lq)[1],2)`, $\beta=$ `r round(coef(model_lq)[2],2)`, $D_0=$ `r round(coef(model_mt)[1],2)`, and n=`r round(coef(model_mt)[2],2)`.
