---
title: "Exploratory Correlation Analysis of Redox Related Metabolites in Erlotinib-treated Head and Neck Cancer Patients"
output: slidy_presentation
---

## Data collected on each patient

- Erlotinib, OSI-420 (oxidated form of erlotinib), and ratio of the two
- Erlotinib dose (smoker 300, non-smoker 150)
- Free, bound, and total cholesterol before and after erlotinib treatment
- Various oxisterols before and after erlotinib treatment
- Cysteine and GSH redox couples before and after erlotinib treatment
- Tumor decrease in response to the treatment
- Free, bound, and total cholesterol before and after erlotinib treatment in buccal swab samples

```{r load data, include=FALSE}
library(readxl)
library(GGally)
library(ggplot2)
url <- "https://github.com/xfreppihs/lab_data_analysis/blob/master/correlation_analysis/erlotinib.xlsx?raw=true"
destfile <- "erlotinib.xlsx"
curl::curl_download(url, destfile)
data <- read_excel(destfile)

#rename column names
colnames(data)<-c("Patient","OSI","Erlotinib","OtoE","FreeCholB","FreeCholA","TotalCholB","TotalCholA","BoundCholB","BoundCholA","DoseofE","Oxy27OHB","Oxy27OHA","Oxy7aOH27OHB","Oxy7aOH27OHA","Oxy7KetoB","Oxy7KetoA","Oxy25OHB","Oxy25OHA","Oxy24OHB","Oxy24OHA","Oxy7aOH25OHCholB","Oxy7aOH25OHCholA","Oxy7aOHB","Oxy7aOHA","Oxy7bOHB","Oxy7bOHA","CysRtoOB","CysRtoOA","GSHRtoOB","GSHRtoOA","TumorDecrease","BuccalTotalCholA","BuccalTotalCholB","BuccalFreeCholA","BuccalFreeCholB","BuccalBoundCholA","BuccalBoundCholB")

#create factor column for DoseofE
data$DoseFactor<-factor(data$DoseofE, labels = c("low dose","high dose"))
```

## erlotinib vs tumor response

```{r erlotinib vs tumor response, echo=FALSE}
a<-ggpairs(data, columns = c("OSI","Erlotinib","OtoE","TumorDecrease"),
           switch = "both", #switch labels to bottom and left
           mapping = aes(color = DoseFactor),
           upper = list(continuous = wrap("cor",size = 3)),
           lower = list(continuous = wrap("points",size= 3)),
           diag = list(continuous = wrap("densityDiag",alpha = 0.5)),
           columnLabels = c("OSI-420","Erlotinib","Oxidized erlotinib%","Tumor decrease"))
a
```

- Tumor decrease is bigger in the high dose group (panel 4.4) and correlates positively with erlotinib (panel 2.4).

## erlotinib vs tumor response

```{r erlotinib, echo=FALSE, fig.show='hold', warning=FALSE, out.width="50%"}
a[4,2]+xlab(expression(paste("Erlotinib (pg/",mu,"L of plasma)")))+
  ylab("Percentage of tumor decrease")+
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 18),
        legend.position=c(0.2,0.9),
        legend.background = element_rect(color = "Black"),
        panel.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        axis.line = element_line(color = "Black"),
        axis.text.x = element_text(color = "Black", size=12),
        axis.text.y = element_text(color = "Black", size=12),
        axis.title.y = element_text(size=18),
        axis.title.x = element_text(size=18))

#Discretize Erlotinib concentration
data$ErlotinibFactor[data$Erlotinib<2000]<-"low Erl"
data$ErlotinibFactor[data$Erlotinib>2000]<-"high Erl"
ggplot(data, aes(x=ErlotinibFactor, y=TumorDecrease))+
  geom_point(size = 4)+geom_boxplot(alpha=0.5)+
  ylab("Percentage of tumor decrease")+
  scale_x_discrete(limits = c("low Erl","high Erl"))+
  theme(panel.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        axis.line = element_line(color="Black"),
        axis.text.x = element_text(color="Black", size=18),
        axis.text.y = element_text(color="Black", size=12),
        axis.title.y = element_text(size=18),
        axis.title.x = element_blank())
```

```{r erlotinib correlation, include=FALSE}
#to get p-values and correlation coefficients, use cor.test
cor.test(data$Erlotinib,data$TumorDecrease)
data_high_dose<-data[data$DoseofE == 300,]
data_low_dose<-data[data$DoseofE == 150,]
cor.test(data_high_dose$Erlotinib,data_high_dose$TumorDecrease)
cor.test(data_low_dose$Erlotinib,data_low_dose$TumorDecrease)

#is TumorDecrease normally distributed?
plot(density(data$TumorDecrease))
qqnorm(data$TumorDecrease)
shapiro.test(data$TumorDecrease)
#p = 0.3661 fail to reject null hypothesis that TumorDecrease is normal

#t-test for TumorDecrease between high/low Erl groups
lowErl = unlist(data[data$ErlotinibFactor == "low Erl","TumorDecrease"])
highErl = unlist(data[data$ErlotinibFactor == "high Erl","TumorDecrease"])
t.test(lowErl,highErl)

#Mann-Whitney U test
wilcox.test(lowErl,highErl)
```

(left) Since tumor response correlates with erlotinib, single out erlotinib vs tumor pair to plot. The correlation coefficient is `r round(cor.test(data$Erlotinib,data$TumorDecrease)[["estimate"]][["cor"]],3)`, with `r round(cor.test(data_high_dose$Erlotinib,data_high_dose$TumorDecrease)[["estimate"]][["cor"]],3)` for the high dose group and `r round(cor.test(data_low_dose$Erlotinib,data_low_dose$TumorDecrease)[["estimate"]][["cor"]],3)` for the low dose group.

(right) Discretize erlotinib concentration into two categories (less or greater than 2000). It is clear that the tumor decrease is higher in the high erlotinib group. p-value = `r round(t.test(lowErl,highErl)[["p.value"]],3)` by t-test, p-value = `r round(wilcox.test(lowErl,highErl)[["p.value"]],3)` by Mann-Whitney U test.

## cholesterol (before) vs tumor response

```{r cholesterol before vs tumor response, echo=FALSE}
ggpairs(data, columns = c("FreeCholB","TotalCholB","BoundCholB","TumorDecrease"),
        switch = "both",
        mapping = aes(color = DoseFactor),
        upper = list(continuous = wrap("cor",size = 3)),
        lower=list(continuous = wrap("points",size= 3)),
        diag = list(continuous = wrap("densityDiag",alpha = 0.5)),
        columnLabels = c("Free cholesterol","Total cholesterol","Bound cholesterol","Tumor decrease"))
```

- Tumor decrease negatively correlates with bound and total cholesterol in patients receiving high dose treatment (panel 2.4, 3.4): maybe related to smoking? This is before treatment, certainly has nothing to do with erlotinib dose.

## cholesterol (after) vs tumor response

```{r cholesterol after vs tumor response, echo=FALSE}
ggpairs(data, columns = c("FreeCholA","TotalCholA","BoundCholA","TumorDecrease"),
        switch = "both",
        mapping = aes(color = DoseFactor),
        upper = list(continuous = wrap("cor",size = 3)),
        lower=list(continuous = wrap("points",size= 3)),
        diag = list(continuous = wrap("densityDiag",alpha = 0.5)),
        columnLabels = c("Free cholesterol","Total cholesterol","Bound cholesterol","Tumor decrease"))
```

- The effect described in previous page is absent after treatment.

## oxysterols (before) vs tumor response

```{r oxysterol before vs tumor response, echo=FALSE}
ggpairs(data, columns = c("Oxy27OHB","Oxy7aOH27OHB","Oxy7aOHB","TumorDecrease"),
        switch = "both",
        mapping = aes(color = DoseFactor),
        upper = list(continuous = wrap("cor",size = 3)),
        lower=list(continuous = wrap("points",size= 3)),
        diag = list(continuous = wrap("densityDiag",alpha = 0.5)),
        columnLabels = c("27-OH","7a-OH-27-OH","7a-OH",
                         "Tumor decrease"))
```

- 7a-OH-27-OH is a really good indicator of tumor response (panel 2.4), so is 27-OH in low dose patients (non-smoker)(panel 1.4).

## oxysterols (after) vs tumor response

```{r oxysterol after vs tumor response, echo=FALSE}
ggpairs(data, columns = c("Oxy27OHA","Oxy7aOH27OHA","Oxy7aOHA","TumorDecrease"),
        switch = "both",
        mapping = aes(color = DoseFactor),
        upper = list(continuous = wrap("cor",size = 3)),
        lower=list(continuous = wrap("points",size= 3)),
        diag = list(continuous = wrap("densityDiag",alpha = 0.5)),
        columnLabels = c("27-OH","7a-OH-27-OH","7a-OH","Tumor decrease"))
```

- After treatment, 7a-OH-27-OH is positively, 7a-OH is negatively correlated with tumor decrease in high dose patients (panel 3.4, 4.4). 

## redox pairs (before) vs tumor response

```{r redox pairs before vs tumor response, echo=FALSE}
ggpairs(data, columns = c("CysRtoOB","GSHRtoOB","TumorDecrease"),
        switch = "both",
        mapping = aes(color = DoseFactor),
        upper = list(continuous = wrap("cor",size = 3)),
        lower=list(continuous = wrap("points",size= 3)),
        diag = list(continuous = wrap("densityDiag",alpha = 0.5)),
        columnLabels = c("Cys+MSBT/Cystine","GSH+MSBT/GSSG",
                         "Tumor decrease"))
```

- Smokers have higher Cys+MSBT/Cystine ratio (panel 1.1).

## redox pairs (after) vs tumor response

```{r redox pairs after vs tumor response, echo=FALSE}
ggpairs(data, columns = c("CysRtoOA","GSHRtoOA","TumorDecrease"),
        switch = "both",
        mapping = aes(color = DoseFactor),
        upper = list(continuous = wrap("cor",size = 3)),
        lower=list(continuous = wrap("points",size= 3)),
        diag = list(continuous = wrap("densityDiag",alpha = 0.5)),
        columnLabels = c("Cys+MSBT/Cystine","GSH+MSBT/GSSG",
                         "Tumor decrease"))
```

- After the treatment, smokers still have higher Cys+MSBT/Cystine ratio (panel 1.1).

```{r remove low GSSG, include=FALSE}
#Remove some low GSSG peaks
GSHRtoOBnew<-data$GSHRtoOB
GSHRtoOAnew<-data$GSHRtoOA
GSHRtoOBnew[c(3,16)]<-NA
GSHRtoOAnew[c(3,16)]<-NA
data<-cbind(data,GSHRtoOBnew,GSHRtoOAnew)
```

## redox pairs (before) vs tumor response, low GSSG points removed

```{r new redox pairs before vs tumor response, echo=FALSE, warning=FALSE}
#plot redox pairs again
ggpairs(data, columns = c("CysRtoOB","GSHRtoOBnew","TumorDecrease"), 
        switch = "both", 
        mapping = aes(color = DoseFactor),
        upper = list(continuous = wrap("cor",size = 3)),
        lower=list(continuous = wrap("points",size= 3)),
        diag = list(continuous = wrap("densityDiag",alpha = 0.5)),
        columnLabels = c("Cys+MSBT/Cystine","GSH+MSBT/GSSG",
                         "Tumor decrease"))
```

- Removed a few really low GSSG data points.

## redox pairs (after) vs tumor response, low GSSG points removed

```{r new redox pairs after vs tumor response, echo=FALSE, warning=FALSE}
ggpairs(data, columns = c("CysRtoOA","GSHRtoOAnew","TumorDecrease"), 
        switch = "both",
        mapping = aes(color = DoseFactor),
        upper = list(continuous = wrap("cor",size = 3)),
        lower=list(continuous = wrap("points",size= 3)),
        diag = list(continuous = wrap("densityDiag",alpha = 0.5)),
        columnLabels = c("Cys+MSBT/Cystine","GSH+MSBT/GSSG",
                         "Tumor decrease"))
```

- Tumor response is inversely correlated with GSH+MSBT/GSSG ratio post-treatment especially in the high erlotinib dose subgroup (panel 2.3). 

## Cys redox vs erlotinib dose (smoking)

```{r cys redox, echo=FALSE, fig.show='hold', warning=FALSE, out.width="50%"}
ggplot(data, aes(x=DoseFactor,y=CysRtoOB))+
  geom_point(size = 5)+geom_boxplot(alpha=0.5)+
  ylab("Cys+MSBT/Cystine (before treatment)")+
  theme(panel.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        axis.line = element_line(color = "Black"),
        axis.text.x = element_text(color = "Black", size=18),
        axis.text.y = element_text(color = "Black", size=12),
        axis.title.y = element_text(size=18),
        axis.title.x = element_blank())
ggplot(data, aes(x=DoseFactor,y=CysRtoOA))+
  geom_point(size = 5)+geom_boxplot(alpha=0.5)+
  ylab("Cys+MSBT/Cystine (after treatment)")+
  theme(panel.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        axis.line = element_line(color = "Black"),
        axis.text.x = element_text(color="Black", size=18),
        axis.text.y = element_text(color="Black", size=12),
        axis.title.y = element_text(size=18),
        axis.title.x = element_blank())
```

```{r Cys stats, include=FALSE}
#are Cys redox values normally distributed?
plot(density(data$CysRtoOB))
plot(density(data$CysRtoOA))
qqnorm(data$CysRtoOB)
qqnorm(data$CysRtoOA)
shapiro.test(data$CysRtoOB)
shapiro.test(data$CysRtoOA)
#Assume normality

#t-test for Cys redox values between high/low dose groups
low_dose_pre = unlist(data[data$DoseFactor == "low dose","CysRtoOB"])
low_dose_post = unlist(data[data$DoseFactor == "low dose","CysRtoOA"])
high_dose_pre = unlist(data[data$DoseFactor == "high dose","CysRtoOB"])
high_dose_post = unlist(data[data$DoseFactor == "high dose","CysRtoOA"])
t.test(low_dose_pre,high_dose_pre)
t.test(low_dose_post,high_dose_post)

#Mann-Whitney U test
wilcox.test(low_dose_pre,high_dose_pre)
wilcox.test(low_dose_post,high_dose_post)
```

- High dose patients (smokers) have higher reduced to oxidized cysteine ratio pre- and post- treatment. (left, before treatment) p-value = `r round(t.test(low_dose_pre,high_dose_pre)[["p.value"]],3)` by t-test, p-value = `r round(wilcox.test(low_dose_pre,high_dose_pre)[["p.value"]],3)` by Mann-Whitney U test. (right, after treatment) p-value = `r round(t.test(low_dose_post,high_dose_post)[["p.value"]],3)` by t-test, p-value = `r round(wilcox.test(low_dose_post,high_dose_post)[["p.value"]],3)` by Mann-Whitney U test.

## Buccal cholesterol (before) vs tumor response

```{r buccal cholesterol before vs tumor response, echo=FALSE, warning=FALSE}
ggpairs(data, columns = c("BuccalFreeCholB","BuccalTotalCholB","BuccalBoundCholB","TumorDecrease"),
        switch = "both",
        mapping = aes(color = DoseFactor),
        upper = list(continuous = wrap("cor",size = 3)),
        lower=list(continuous = wrap("points",size= 3)),
        diag = list(continuous = wrap("densityDiag",alpha = 0.5)),
        columnLabels = c("Free cholesterol","Total cholesterol","Bound cholesterol", "Tumor decrease"))
```

## Buccal cholesterol (after) vs tumor response

```{r buccal cholesterol after vs tumor response, echo=FALSE, warning=FALSE}
ggpairs(data, columns = c("BuccalFreeCholA","BuccalTotalCholA","BuccalBoundCholA","TumorDecrease"),
        switch = "both",
        mapping = aes(color = DoseFactor),
        upper = list(continuous = wrap("cor",size = 3)),
        lower=list(continuous = wrap("points",size= 3)),
        diag = list(continuous = wrap("densityDiag",alpha = 0.5)),
        columnLabels = c("Free cholesterol","Total cholesterol","Bound cholesterol", "Tumor decrease"))
```

- Total and bound cholesterol in post-treatment buccal swab samples correlates with tumor decrease especially in the high erlotinib dose group (panel 2.4, 3.4). Keep in mind that bound cholesterol has some negative values (from total - free), so the accuracy of measurement is questionable.

## plasma cholesterol before and after treatment

```{r cholesterol, echo=FALSE}
#plot for each patients the plasma cholesterol (free, bound, and total) before and after treatment
#subset plasma cholesterol data, melt it from wide to long
library(reshape2)
data_pchol_wide<-data[,c(1,5:10)]
data_pchol_long<-melt(data_pchol_wide, id.vars=c("Patient"))

data_pchol_long$time<-NA #creating a new column for time
data_pchol_long$time[grep("B$",data_pchol_long$variable)]<-"Before" #anything that ends with B means before treatment
data_pchol_long$time[grep("A$",data_pchol_long$variable)]<-"After"

data_pchol_long$cholesterol<-NA #creating a new column for cholesterol type
data_pchol_long$cholesterol[grep("Free",data_pchol_long$variable)]<-"Free"
data_pchol_long$cholesterol[grep("Bound",data_pchol_long$variable)]<-"Bound" 
data_pchol_long$cholesterol[grep("Total",data_pchol_long$variable)]<-"Total" 

ggplot(data_pchol_long,aes(x=time,y=value))+
  geom_line(aes(group=Patient,color=as.character(Patient)),size=1)+
  facet_grid(.~cholesterol)+labs(color = "Patient")+
  scale_x_discrete(limits=c("Before","After"))+ #changing the order of a discrete x scale
  ylab(expression(paste("Cholesterol (",mu,"g/",mu,"L of plasma)")))+
  theme(panel.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.background = element_rect(color = "black"),#legend border
        axis.line = element_line(color = "Black"),#axis line
        axis.text.x = element_text(color="Black", size=10),
        axis.text.y = element_text(color="Black", angle = 90, hjust=0.5,
                                   size=10),
        axis.title.y = element_text(size=10),
        axis.title.x = element_blank())
```

- Interestingly, plasma cholesterol seems to be increased in most patients after treatment.

## Multiple linear regression - single predictor
We want to build a model using multiple variables to predict tumor response. For model selection, here we use the forward selection with adjusted R^2 method. We choose single predictors from the plasma samples before the treatment.

- erlotinib adjusted R^2 = `r summary(lm(TumorDecrease~Erlotinib,data))$adj.r.squared`
- total cholesterol adjusted R^2 = `r summary(lm(TumorDecrease~TotalCholB,data))$adj.r.squared`
- 7a-OH-27-OH adjusted R^2 = `r summary(lm(TumorDecrease~Oxy7aOH27OHB,data))$adj.r.squared`
- GSH-MSBT/GSSG ratio adjusted R^2 = `r summary(lm(TumorDecrease~GSHRtoOBnew,data))$adj.r.squared`

So first we pick 7a-OH-27-OH.

## Multiple linear regression - adding second predictor

- 7a-OH-27-OH + erlotinib adjusted R^2 = `r summary(lm(TumorDecrease~Oxy7aOH27OHB+Erlotinib,data))$adj.r.squared`
- 7a-OH-27-OH + total cholesterol adjusted R^2 = `r summary(lm(TumorDecrease~Oxy7aOH27OHB+TotalCholB,data))$adj.r.squared`
- 7a-OH-27-OH + GSH-MSBT/GSSG ratio adjusted R^2 = `r summary(lm(TumorDecrease~Oxy7aOH27OHB+GSHRtoOBnew,data))$adj.r.squared`

So we go with 7a-OH-27-OH + erlotinib.

## Multiple linear regression - adding third predictor

- 7a-OH-27-OH + erlotinib + total cholesterol adjusted R^2 = `r summary(lm(TumorDecrease~Oxy7aOH27OHB+Erlotinib+TotalCholB,data))$adj.r.squared`
- 7a-OH-27-OH + erlotinib + GSH-MSBT/GSSG ratio adjusted R^2 = `r summary(lm(TumorDecrease~Oxy7aOH27OHB+Erlotinib+GSHRtoOBnew,data))$adj.r.squared`

So we go with 7a-OH-27-OH + erlotinib + GSH-MSBT/GSSG ratio.

## Multiple linear regression - final model

- 7a-OH-27-OH + erlotinib + GSH-MSBT/GSSG ratio + total cholesterol adjusted R^2 = `r summary(lm(TumorDecrease~Oxy7aOH27OHB+Erlotinib+GSHRtoOBnew+TotalCholB,data))$adj.r.squared`
- Adding total cholesterol doesn't help, so we drop it to get the final model.

```{r mlr, echo=FALSE}
mlr<-lm(TumorDecrease~Oxy7aOH27OHB+Erlotinib+GSHRtoOBnew,data)
summary(mlr)
```